---
title: "ESVCP 2022: Pre-congress Session: Parametric versus Non-parametric statistics"
output:
    html_document:
        code_download: false
        toc: false
        number_sections: false
        code_folding: "none"
---


```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

***

```{r out.width='80%',fig.asp=.8, fig.align='center',echo=FALSE}
if ("pi" %in% ls()) rm("pi")
kopvoeter <- function(x, y, angle = 0, l = .2, cex.dot = .5, pch = 19, col = "black") {
  angle <- angle / 180 * pi
  points(x, y, cex = cex.dot, pch = pch, col = col)
  lines(c(x, x + l * cos(-pi / 2 + angle)), c(y, y + l * sin(-pi / 2 + angle)), col = col)
  lines(c(x + l / 2 * cos(-pi / 2 + angle), x + l / 2 * cos(-pi / 2 + angle) + l / 4 * cos(angle)), c(y + l / 2 * sin(-pi / 2 + angle), y + l / 2 * sin(-pi / 2 + angle) + l / 4 * sin(angle)), col = col)
  lines(c(x + l / 2 * cos(-pi / 2 + angle), x + l / 2 * cos(-pi / 2 + angle) + l / 4 * cos(pi + angle)), c(y + l / 2 * sin(-pi / 2 + angle), y + l / 2 * sin(-pi / 2 + angle) + l / 4 * sin(pi + angle)), col = col)
  lines(c(x + l * cos(-pi / 2 + angle), x + l * cos(-pi / 2 + angle) + l / 2 * cos(-pi / 2 + pi / 4 + angle)), c(y + l * sin(-pi / 2 + angle), y + l * sin(-pi / 2 + angle) + l / 2 * sin(-pi / 2 + pi / 4 + angle)), col = col)
  lines(c(x + l * cos(-pi / 2 + angle), x + l * cos(-pi / 2 + angle) + l / 2 * cos(-pi / 2 - pi / 4 + angle)), c(y + l * sin(-pi / 2 + angle), y + l * sin(-pi / 2 + angle) + l / 2 * sin(-pi / 2 - pi / 4 + angle)), col = col)
}

par(mar = c(0, 0, 0, 0), mai = c(0, 0, 0, 0))
plot(0, 0, xlab = "", ylab = "", xlim = c(0, 10), ylim = c(0, 10), col = 0, xaxt = "none", yaxt = "none", axes = FALSE)
rect(0, 6, 10, 10, border = "red", lwd = 2)
text(.5, 8, "population", srt = 90, col = "red", cex = 2)
symbols(3, 8, circles = 1.5, col = "red", add = TRUE, fg = "red", inches = FALSE, lwd = 2)
set.seed(330)
grid <- seq(0, 1.3, .01)

for (i in 1:50)
{
  angle1 <- runif(n = 1, min = 0, max = 360)
  angle2 <- runif(n = 1, min = 0, max = 360)
  radius <- sample(grid, prob = grid^2 * pi / sum(grid^2 * pi), size = 1)
  kopvoeter(3 + radius * cos(angle1 / 180 * pi), 8 + radius * sin(angle1 / 180 * pi), angle = angle2)
}
text(7.5, 8, "Cholesterol in population", col = "red", cex = 1.2)

rect(0, 0, 10, 4, border = "blue", lwd = 2)
text(.5, 2, "sample", srt = 90, col = "blue", cex = 2)
symbols(3, 2, circles = 1.5, col = "red", add = TRUE, fg = "blue", inches = FALSE, lwd = 2)
for (i in 0:2) {
  for (j in 0:4)
  {
    kopvoeter(2.1 + j * (3.9 - 2.1) / 4, 1.1 + i)
  }
}
text(7.5, 2, "Cholesterol in sample", col = "blue", cex = 1.2)

arrows(3, 5.9, 3, 4.1, col = "black", lwd = 3)
arrows(7, 4.1, 7, 5.9, col = "black", lwd = 3)
text(1.5, 5, "EXP. DESIGN (1)", col = "black", cex = 1.2)
text(8.5, 5, "ESTIMATION &\nINFERENCE (3)", col = "black", cex = 1.2)
text(7.5, .5, "DATA EXPLORATION &\nDESCRIPTIVE STATISTICS (2)", col = "black", cex = 1.2)
```

## Course Description

This session covers an introduction to parametric and nonparametric hypothesis testing by examples. 



---

## Software

- The learning materials are developed for [R](https://www.r-project.org/) version 4.1.0 or later. We also recommend to also install the latest version of [RStudio](https://www.rstudio.com/products/rstudio/download/).

- To install all requirements, please copy and paste this lines of code in your R console:

```r
source("https://raw.githubusercontent.com/statOmics/PSLS/master/install.R")
```

- To get familiar with most important concepts in working with R and Rmarkdown, you may download the respective cheat sheets:

    - [Base R](https://iqss.github.io/dss-workshops/R/Rintro/base-r-cheat-sheet.pdf)
    - [Rmarkdown](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
    - [Other](https://www.rstudio.com/resources/cheatsheets/)

## Learning materials

1. [Introduction](./01-intro.html)
2. [Rank Tests](./02-ranktests.html)

## Instructor

 - [Lieven Clement](./instructors.html)